---
title: "Medical Spa Product Projections"
author: "Cameron Houston"
date: "2025-01-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(xgboost)
library(fpp3)
library(fable)
library(tsibble)
library(RColorBrewer)
```

```{r}
#Import Data 
sales <- read.csv("2022-2024 Clinic Product Sales Data.csv")
promos <- read.csv("Promotion details and dates.csv")
```


## Cleaning of Data 
```{r}
#Make Dates Correct Format 
promos_clean <- promos %>% mutate(Start.Date = as.Date(Start.Date, format = "%m/%d/%Y"), 
                                  End.Date = as.Date(End.Date, format = "%m/%d/%Y"))

sales_promo <- sales %>% mutate(sale.date = as.Date(sale.date, format = "%m/%d/%Y")) %>% 
  select(-guest.user.id)

##Add Promotion by Date to Sales to Explore 
#Promotion Column for No Promotion
sales_promo$promo <- 0
#Promotion Column for Type
sales_promo$promotion.type <- "None"
#For Loop of Each Promotion
for(i in seq_len(nrow(promos_clean))){
  promotion <- promos_clean[i,]
  promotion_name <- promotion[1]
  #Promotion Dates
  sales_promo$promo <- ifelse(sales_promo$sale.date >= promotion$Start.Date
                        & sales_promo$sale.date <= promotion$End.Date,1, sales_promo$promo)
  #Promotion Type
  sales_promo$promotion.type <- ifelse(sales_promo$promo == 1 & sales_promo$promotion.type == "None", promotion_name, sales_promo$promotion.type)
  sales_promo$promotion.type <- unlist(sales_promo$promotion.type)
}
```



##Exploring Sales
```{r}
# All Product Sales 
##Explore the sales over years
sales_promo %>% 
  separate(sale.date, into=c("year", "month", "day"),sep = "-") %>%
  group_by(year) %>% 
  summarise(yearly.sales = sum(quantity.sold)) %>% 
  ggplot(aes(x = year, y = yearly.sales)) + geom_col()+ 
  labs(title = "All Product Yearly Sales")

#Serum Product Sales
##Explore hyaluronic sales over years
sales_promo %>% 
  separate(sale.date, into=c("year", "month", "day"),sep = "-") %>%
  filter(product.name == "Hyaluronic Acid Serum") %>% 
  group_by(year) %>% 
  summarise(yearly.sales = sum(quantity.sold)) %>% 
  ggplot(aes(x = year, y = yearly.sales)) + 
  geom_col() + 
  labs(title = "Hyaluronic Acid Serum Yearly Sales")

#Serum Promotions
##Explore types of promotions and sales over years
sales_promo %>% 
  separate(sale.date, into=c("year", "month", "day"), sep = "-") %>% 
  filter(product.name == "Hyaluronic Acid Serum") %>% 
  group_by(promotion.type, year) %>% 
  summarise(serum.sold = sum(quantity.sold))

#Clinic G Serum Sales

##Product sales by promotion types
sales_promo %>% 
  separate(sale.date, into=c("year", "month", "day"), sep = "-") %>% 
  filter(product.name == "Hyaluronic Acid Serum" & clinic.name == "Clinic G") %>% 
  group_by(promotion.type, year) %>% 
  summarise(serum.sold = sum(quantity.sold)) %>% 
  arrange(desc(serum.sold))

##What other products are popular at clinic g over the years
sales_promo %>% 
  separate(sale.date, into=c("year", "month", "day"), sep = "-") %>% 
  filter(clinic.name == "Clinic G") %>% 
  group_by(product.name, year) %>% 
  summarise(products.sold = sum(quantity.sold)) %>% 
  arrange(desc(products.sold))

#Hyaluronic acid serum sold at clinic g over the years
sales_promo %>% 
  separate(sale.date, into=c("year", "month", "day"), sep = "-") %>% 
  filter(product.name == "Hyaluronic Acid Serum" & clinic.name == "Clinic G") %>% 
  group_by(promotion.type, year) %>% 
  summarise(serum.sold = sum(quantity.sold)) %>% 
  ggplot(aes(x = year, y = serum.sold)) + 
  geom_col() + 
  labs(title = "Hyaluronic Acid Serum Yearly Sales at Clinic G")
```
Hyaluronic Acid Serum sales are up over the years, including at Clinic G. Only serums were included in the data set, so comparing the type of product to the promotion are not needed in this dataset. Although, the serum promotion has higher serum sales than any other promotion. 


## Modeling 
Start with the basics for quarterly product sold, then can evaluate additional components such as promotions
```{r}
##Filter Product and Clinic
sales_clean <- sales_promo %>% filter(product.name == "Hyaluronic Acid Serum", clinic.name == "Clinic G")
#Make Tsibble for Month (will aggregate for quarter later due to sample size)
sales_ts <- sales_clean %>% mutate(Month = yearmonth(sale.date)) %>% group_by(Month) %>%
  summarize(total.sold = sum(quantity.sold)) %>% as_tsibble(index = Month)
##Split test/train
train <- sales_ts %>% 
  filter_index(~ "2024-06")
test <- sales_ts %>%
  filter_index("2024-07" ~ "2024-09") #Quarter 3 
```

```{r}
#Explore time series relationship 
autoplot(train, total.sold)
train %>% features(total.sold, feat_stl)
```
Ft = 0.33
Fs = 0.41
Indicating low to moderate strengths of trend and seasons, will explore further

```{r}
#Seasonal Trend Losses Decomposition
dcmp <- train %>%  model(stl = STL(total.sold))

#STL decomp visuals 
components(dcmp) %>% as_tsibble() |>
autoplot(total.sold, colour="black") +
geom_line(aes(y=trend), colour = "blue") + labs( y =
"Total Products Sold", title = "Hyaluronic Serum Sold by Month" )
components(dcmp) %>%  as_tsibble() %>% 
autoplot(total.sold, colour="black") +
geom_line(aes(y=season_adjust), colour = "blue") + labs( y =
"Total Products Sold", title = "Hyaluronic Serum Sold by Month Seasonally Adjusted" )
#Visually look at seasons and trends
components(dcmp) %>% autoplot()
```

```{r}
#Stationary Test 
train %>% features(total.sold, unitroot_kpss) ## 0.0722 -- failing to reject 
train %>%  features(total.sold, unitroot_ndiffs)# Double checking test, 0 differences
##Visually look before continuing on 
train %>% gg_tsdisplay(total.sold, plot_type = "partial")
```
The p value is 0.08 and seasonal differences are not recommended. Will fail to reject the null, favoring that the data is stationary. Will explore models with no transformations. 


```{r}
##Explore Forecast Models 
sales_fit <- train |>
  model(
    SES = ETS(total.sold ~ error("A") + trend("N") + season("N")), #Single exponential smoothing
    `Linear` = ETS(total.sold ~ error("A") + trend("A") + season("N")), #Linear
    `Damped Linear` = ETS(total.sold ~ error("A") + trend("Ad") + season("N")), #Damped linear
    HWAdd = ETS(total.sold ~ error("A") + trend("A") + season("A")), #Holt winters additive
    HWMult = ETS(total.sold ~ error("M") + trend("A") + season("M")) #Holt winters multiplicative
  )
#Evaluate all models
report(sales_fit)
##Evaluate ARIMAS
sales_model <- train %>% model(
 ar3= ARIMA(total.sold ~ pdq(3,0,0)),#based off pacf graph
 ma2 = ARIMA(total.sold ~ pdq(0,0,2)), #based off acf graph
 rand1 = ARIMA(total.sold)) # using automated search

#Evaluate all models
report(sales_model)
sales_model %>% select(rand1) %>% report #Check automated model
```
Top models by AICc: ARIMAs --> Moving Average(2) and ARIMA(0,0,0)
Other Models --> Simple exponential smoothing


```{r}
#Step Intervention
##Consider when promotion exists
#Requires more granular level, reset tsibble for days rather than months
promos_ts <- sales_promo %>% 
  filter(clinic.name =="Clinic G" & product.name == "Hyaluronic Acid Serum") %>% 
  group_by(sale.date, promo, promotion.type) %>%  
  summarize(total.sold = sum(quantity.sold), .groups = "drop") %>% as_tsibble(index = sale.date)
##New split test/train
train2 <- promos_ts %>% 
  filter_index(~ "2024-06")
test2 <- promos_ts %>%
  filter_index("2024-07" ~ "2024-09")
```


```{r}
#Check for missing days 
count_gaps(train2)
# Train needs to go until day before test, fill in last days
last_date <- max(train2$sale.date)  
target_end_date <- as.Date("2024-06-30")
# Generate missing dates
missing_dates <- seq(last_date + 1, target_end_date, by = "day")
# Missing values with imputed values
missing_data <- tibble(
  sale.date = missing_dates,
  promo = 0, 
  total.sold = 0, 
  promotion.type = "None"
)

# Combine the original data with the missing data
train2<- bind_rows(train2, missing_data)
#Fill in gaps from original 
train2 <- fill_gaps(train2, promo = 0, total.sold = 0, promotion.type = "None") 
#Repeat for potential missing test dates
test2 <- fill_gaps(test2, promo = 0, total.sold = 0, promotion.type = "None")
```
 
```{r}
#Step Intervention Model
step_model_promo <- train2 %>% 
  model(mod1 = ARIMA(total.sold ~ pdq(0,0,1)+promo), ##Add promotion as additional var
  mod2 = ARIMA(total.sold ~ pdq(0,0,2)), 
  rand1 = ARIMA(total.sold))
report(step_model_promo)

```

 

```{r}
#Try all models explored
top_mods <- train %>% 
  model(mod1 = ARIMA(total.sold ~ pdq(0,0,2)), 
        mod2 = ARIMA(total.sold ~ pdq(1,0,2)), 
        mod3 = ARIMA(total.sold ~ pdq(0,0,0)),
        SES = ETS(total.sold ~ error("A") + trend("N") + season("N")))
report(top_mods)
#Forecast models
top_fore <- forecast(top_mods, test)
#Plot forecast 
top_fore %>%
  autoplot(train) + 
  autolayer(fitted(top_mods), col = "blue", linetype = "dashed") + 
  ylab("Product Sold")
  
## Check accuracy / model evaluation
fabletools::accuracy(top_fore, test)
```


```{r}
#Simple Average Ensemble Model
##Ensemble the arima models, as they have shown to be best
ensemble_mod <- train %>% 
  model(mod1 = ARIMA(total.sold ~ pdq(0,0,2)), 
        mod2 = ARIMA(total.sold ~ pdq(1,0,2)),
        mod3 = ARIMA(total.sold ~ pdq(1,0,2))) %>% 
  mutate(combo = (mod1 + mod2 + mod3) / 3) #Simple average of all models
#Forecast ensemble model
model_all_for <- ensemble_mod %>%
  forecast(h = 3)

##Visualize all predictions
model_all_for %>%
  autoplot() + 
  scale_colour_brewer(palette = "Paired") +
  labs(y = "Products Sold", title = "Comparison of Forecasts") 
## Get ensemble evaluations
model_all_for %>%
  accuracy(test) %>%
  arrange(MAPE)
```
While the ensemble model did perform better than most models alone, model one, the 2 moving average lags performed the overall best. 


##Selected Model 
Best model overall was MA(2), now going to evaluate final predictions`
```{r}
#Run model alone
demand_model <- train %>% 
  model(mod1 = ARIMA(total.sold ~ pdq(0,0,2))) 
#Pull single forecast
demand_forecast <- forecast(demand_model, h = 3)
```

## Final Prediction Forecast
```{r}
## Goal was to predict quarter sales, need to aggregate months
#Forecast Predictions
predictions <- demand_forecast$.mean
actual <- test$total.sold
sum(predictions)
MAE <- abs(sum(predictions)-sum(actual))
MAE
```
Final Predictions for Q3 in 2024 were 124 containers of Hyaluronic Acid Serum to be sold at Clinic G. However, I under predicted by 10 containers. I believe additional variables or better understanding of the industry would allow me to tailor this or develop more advanced forecast models that utilize machine learning algorithms like neural network or extreme gradient boosting. 






